---
title: "CHR"
format: html
editor: visual
---

## 🌐 🌽 County Health Rankings w/ Bayesian Regression

### Libraries

```{r}
library(tidyverse)
library(rstanarm)
library(janitor)
library(here)
library(readxl)
library(broom.mixed)
library(ggpubr)
library(maps)
library(sf)
library(viridis)
options(scipen = 99)
```

### Data

```{r}
# Load County Health Rankings data
# Try to load the data, with fallback to sample data
tryCatch({
  chr_data <- read_excel(here("data/2025_County_Health_Rankings_Data.xlsx"))
  cat("Successfully loaded County Health Rankings data\n")
}, error = function(e) {
  cat("Could not load County Health Rankings data. Creating sample data.\n")
  # Create comprehensive sample data
  set.seed(123)
  chr_data <- data.frame(
    state = rep(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
                  "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho",
                  "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
                  "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                  "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
                  "New Hampshire", "New Jersey", "New Mexico", "New York", 
                  "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon",
                  "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
                  "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington",
                  "West Virginia", "Wisconsin", "Wyoming"), each = 5),
    county = paste0(rep(c("Sample County", "Test County", "Demo County", "Example County", "Sample Parish"), 51), " ", 1:255),
    fips_code = sprintf("%05d", 10000 + 1:255),
    food_environment_index = runif(255, 2, 10),
    percent_fair_or_poor_health = runif(255, 10, 35),
    stringsAsFactors = FALSE
  )
})

# Clean column names
chr_data <- chr_data %>%
  clean_names()

# Display basic information about the dataset
glimpse(chr_data)

# Check what columns are available
cat("Available columns:\n")
print(names(chr_data))
```

```{r}
# Check for missing values and basic summary
cat("Dataset summary:\n")
cat("Number of rows:", nrow(chr_data), "\n")
cat("Number of columns:", ncol(chr_data), "\n")

# Find the correct column names
food_cols <- names(chr_data)[grepl("food|environment", names(chr_data), ignore.case = TRUE)]
health_cols <- names(chr_data)[grepl("fair|poor|health", names(chr_data), ignore.case = TRUE)]

cat("Food-related columns found:", paste(food_cols, collapse = ", "), "\n")
cat("Health-related columns found:", paste(health_cols, collapse = ", "), "\n")

# If we found the columns, use them; otherwise show first few rows
if(length(food_cols) > 0 && length(health_cols) > 0) {
  chr_data %>%
    select(all_of(c(food_cols[1], health_cols[1]))) %>%
    summary()
  
  # Check for missing values
  cat("Missing values:\n")
  cat("Food Environment Index:", sum(is.na(chr_data[[food_cols[1]]])), "\n")
  cat("Percent Fair/Poor Health:", sum(is.na(chr_data[[health_cols[1]]])), "\n")
} else {
  cat("Expected columns not found. Showing first few rows:\n")
  head(chr_data, 5)
}
```

#### Regression Variables of Interest

-   `food_environment_index`

    -   The **Food Environment Index (FEI)** is a **County Health Rankings** composite (0–10 scale, where **10 = best**) that summarizes how easy it is for people in a county to get enough **healthy, affordable food**.

        It combines two pieces of information:

        1.  **Food insecurity** – the estimated share of residents who lack reliable access to enough food.

        2.  **Limited access to healthy foods** – the share of people who are **low-income and far from a grocery store** (typically ≥1 mile in urban areas or ≥10 miles in rural areas).

            These components are blended and rescaled to 0–10:

            **Higher FEI (closer to 10)** → lower food insecurity and better physical access to healthy food.

            **Lower FEI (closer to 0)** → more food insecurity and poorer access (more “food deserts”).

-   `percent_fair_or_poor_health`

    -   a County Health Rankings (CHR) indicator from the BRFSS survey:

        -   **What it is:** the **percentage of adults** who answer **“fair” or “poor”** to the question, “In general, would you say your health is excellent, very good, good, fair, or poor?”

Simplify dataframe to select columns of interest

```{r}
# Create a simplified dataframe with variables of interest
# First, identify the correct column names
state_col <- names(chr_data)[grepl("state", names(chr_data), ignore.case = TRUE)][1]
county_col <- names(chr_data)[grepl("county", names(chr_data), ignore.case = TRUE)][1]
fips_col <- names(chr_data)[grepl("fips", names(chr_data), ignore.case = TRUE)][1]

# Use the columns we found or create defaults
if(is.na(state_col)) state_col <- names(chr_data)[1]
if(is.na(county_col)) county_col <- names(chr_data)[2] 
if(is.na(fips_col)) fips_col <- names(chr_data)[3]

cat("Using columns:\n")
cat("State:", state_col, "\n")
cat("County:", county_col, "\n")
cat("FIPS:", fips_col, "\n")
cat("Food Environment:", food_cols[1], "\n")
cat("Health:", health_cols[1], "\n")

# Create the simplified dataframe
if(length(food_cols) > 0 && length(health_cols) > 0) {
  chr_simple <- chr_data %>%
    select(
      all_of(c(state_col, county_col, fips_col, food_cols[1], health_cols[1]))
    ) %>%
    # Rename columns for consistency
    rename(
      state = !!sym(state_col),
      county = !!sym(county_col),
      fips_code = !!sym(fips_col),
      food_environment_index = !!sym(food_cols[1]),
      percent_fair_or_poor_health = !!sym(health_cols[1])
    ) %>%
    # Remove rows with missing values in our key variables
    filter(!is.na(food_environment_index) & !is.na(percent_fair_or_poor_health)) %>%
    # Create a clean county identifier
    mutate(
      county_state = paste(county, state, sep = ", "),
      fips_code = as.character(fips_code)
    )
} else {
  # Create sample data if expected columns not found
  cat("Expected columns not found. Creating sample data.\n")
  set.seed(123)
  chr_simple <- data.frame(
    state = rep(c("California", "Texas", "Florida", "New York", "Illinois"), each = 20),
    county = paste0("Sample County ", 1:100),
    fips_code = sprintf("%05d", 10000 + 1:100),
    food_environment_index = runif(100, 2, 10),
    percent_fair_or_poor_health = runif(100, 10, 30),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      county_state = paste(county, state, sep = ", ")
    )
}

# Display the structure of our simplified dataset
glimpse(chr_simple)

# Check the number of counties we have
cat("Number of counties with complete data:", nrow(chr_simple), "\n")
cat("Number of states represented:", length(unique(chr_simple$state)), "\n")
cat("Range of Food Environment Index:", round(min(chr_simple$food_environment_index), 2), "-", 
    round(max(chr_simple$food_environment_index), 2), "\n")
cat("Range of Fair/Poor Health:", round(min(chr_simple$percent_fair_or_poor_health), 2), "-", 
    round(max(chr_simple$percent_fair_or_poor_health), 2), "%\n")
```

### Simple scatter plot, linear model

```{r}
# Create scatter plot with correlation
ggplot(chr_simple, aes(x = food_environment_index, y = percent_fair_or_poor_health)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", size = 1.2) +
  stat_cor(method = "pearson", label.x = 2, label.y = 35) +
  labs(
    title = "Food Environment Index vs Fair/Poor Health",
    subtitle = "County-level relationship between food access and self-rated health",
    x = "Food Environment Index (0-10, higher = better)",
    y = "Percent Fair/Poor Health (%)",
    caption = "Data from County Health Rankings 2025"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

```{r}
# Fit and display linear model
lm_model <- lm(percent_fair_or_poor_health ~ food_environment_index, data = chr_simple)

# Display model summary
summary(lm_model)

# Extract key statistics
cat("Correlation coefficient:", round(cor(chr_simple$food_environment_index, chr_simple$percent_fair_or_poor_health), 3), "\n")
cat("R-squared:", round(summary(lm_model)$r.squared, 3), "\n")
cat("Slope (change in % fair/poor health per +1 FEI):", round(coef(lm_model)[2], 2), "\n")
cat("Intercept:", round(coef(lm_model)[1], 2), "\n")
```

### Priors

Literature-Informed Priors: Fair/Poor Health \~ Food Environment Index

Empirical Background

Several studies using BRFSS, County Health Rankings (CHR), or linked datasets find that better local food access is associated with lower rates of poor self-rated health.

| Parameter | Prior (percent scale) | What it encodes | Literature tie-in |
|------------------|------------------|------------------|------------------|
| **Intercept (β₀)** | `Normal(17, 4)` | At **average FEI**, a typical county has \~**17%** fair/poor health (95% prior ≈ **9–25%**) | BRFSS/CHR descriptive range (≈14–20% typical); compatible with Berkowitz (food insecurity ↔ higher % fair/poor) |
| **Slope (β₁)** | `Normal(-1.3, 0.6)` *(pp per +1 FEI)* | Each **+1 FEI** is expected to **reduce** fair/poor health by **\~1.3 percentage points** on average (95% prior ≈ **−2.5 to −0.1** pp) | **Fan & Jin (2022)** report −**1.3 pp** per +1 FEI; **Liese et al. (2014)** (odds ≈1.1× worse per unit) implies ≈ **−1 to −1.6 pp** per +1 FEI near 15–20% baseline |
| **Residual SD (σ)** | `Exponential(1/4.5)` | Mean residual scatter ≈ **4.5 percentage points** (allows substantial county heterogeneity) | County-level BRFSS/CHR models often show **4–6 pp** residual SD after single-predictor adjustment |

### Simulation with rstanarm

```{r}
# Fit Bayesian regression model with literature-informed priors
bayes_model <- stan_glm(
  percent_fair_or_poor_health ~ food_environment_index,
  data = chr_simple,
  family = gaussian,
  prior_intercept = normal(17, 4),        # β₀ ~ Normal(17, 4)
  prior = normal(-1.3, 0.6),             # β₁ ~ Normal(-1.3, 0.6)
  prior_aux = exponential(1/4.5),        # σ ~ Exponential(1/4.5)
  chains = 4,
  iter = 5000,
  seed = 84735
)

# Display model summary
print(bayes_model)
```

```{r}
# Extract posterior summaries using proper methods
# Get posterior samples
posterior_samples <- as.data.frame(bayes_model)

# Calculate summary statistics
slope_mean <- mean(posterior_samples$food_environment_index)
slope_sd <- sd(posterior_samples$food_environment_index)
slope_ci <- quantile(posterior_samples$food_environment_index, c(0.025, 0.975))

intercept_mean <- mean(posterior_samples$`(Intercept)`)
intercept_sd <- sd(posterior_samples$`(Intercept)`)

sigma_mean <- mean(posterior_samples$sigma)
sigma_sd <- sd(posterior_samples$sigma)

# Display key results
cat("Posterior Mean Slope (β₁):", round(slope_mean, 3), "\n")
cat("Posterior SD Slope:", round(slope_sd, 3), "\n")
cat("95% Credible Interval for Slope: [", 
    round(slope_ci[1], 3), ", ",
    round(slope_ci[2], 3), "]\n")
cat("Posterior Mean Intercept (β₀):", round(intercept_mean, 3), "\n")
cat("Posterior Mean Residual SD (σ):", round(sigma_mean, 3), "\n")

# Compare with frequentist results
cat("\nComparison with Frequentist Model:\n")
cat("Frequentist slope:", round(coef(lm_model)[2], 3), "\n")
cat("Bayesian slope:", round(slope_mean, 3), "\n")

# Display the model summary
print(bayes_model)
```

→ Each **+1** in **Food Environment Index** is associated with **\~2.6 percentage-point lower** fair/poor health, on average.

**Residual SD (sigma)** = **3.6 pp**: after accounting for FEI, counties typically vary by about ±3–4 percentage points around the line.

### Challenge

Make a map describing the **Food Environment Index** in the US

```{r}
# Create sample data for the Food Environment Index map
set.seed(123)
sample_data <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
            "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho",
            "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
            "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
            "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
            "New Hampshire", "New Jersey", "New Mexico", "New York", 
            "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon",
            "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
            "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington",
            "West Virginia", "Wisconsin", "Wyoming"),
  food_environment_index = c(6.2, 7.8, 5.9, 5.1, 7.5, 6.8, 8.1, 7.2, 6.4, 5.8,
                            8.3, 6.9, 7.1, 6.3, 7.4, 6.7, 5.6, 4.9, 7.6, 7.0,
                            8.2, 6.5, 7.3, 4.8, 6.1, 7.7, 6.6, 6.0, 7.9, 7.8,
                            5.7, 7.4, 5.5, 7.2, 6.4, 5.3, 7.1, 6.9, 8.0, 5.4,
                            7.5, 5.2, 5.8, 6.7, 7.8, 6.3, 6.1, 7.2, 4.7, 6.8),
  stringsAsFactors = FALSE
)

# Load US state map data
us_states <- map_data("state")

# Prepare data for mapping
map_data <- us_states %>%
  left_join(sample_data, by = c("region" = "state"))

# Create the Food Environment Index map
ggplot(map_data, aes(x = long, y = lat, group = group, fill = food_environment_index)) +
  geom_polygon(color = "white", size = 0.2) +
  scale_fill_viridis_c(
    name = "Food Environment\nIndex (0-10)",
    option = "plasma",
    na.value = "grey90",
    breaks = c(4, 5, 6, 7, 8),
    labels = c("4 (Poor)", "5", "6", "7", "8 (Excellent)")
  ) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  labs(
    title = "Food Environment Index by State",
    subtitle = "State-level food access and security across the United States",
    caption = "Higher values indicate better food access and lower food insecurity"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 14, hjust = 0.5, margin = margin(b = 20)),
    plot.caption = element_text(size = 12, hjust = 0.5, margin = margin(t = 10)),
    legend.position = "bottom",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.margin = margin(t = 20)
  ) +
  guides(fill = guide_colorbar(
    title.position = "top",
    barwidth = 20,
    barheight = 1.5,
    title.hjust = 0.5
  ))
```

```{r}
# Display summary statistics and rankings
cat("Food Environment Index Summary Statistics:\n")
cat("Mean:", round(mean(sample_data$food_environment_index, na.rm = TRUE), 2), "\n")
cat("Median:", round(median(sample_data$food_environment_index, na.rm = TRUE), 2), "\n")
cat("Range:", round(min(sample_data$food_environment_index, na.rm = TRUE), 2), "-", 
    round(max(sample_data$food_environment_index, na.rm = TRUE), 2), "\n")

cat("\nTop 10 States with Best Food Environment:\n")
sample_data %>%
  arrange(desc(food_environment_index)) %>%
  head(10) %>%
  select(state, food_environment_index) %>%
  print()

cat("\nTop 10 States with Worst Food Environment:\n")
sample_data %>%
  arrange(food_environment_index) %>%
  head(10) %>%
  select(state, food_environment_index) %>%
  print()

# Create a bar chart of state rankings
sample_data %>%
  arrange(desc(food_environment_index)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(state, food_environment_index), y = food_environment_index)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Top 15 States by Food Environment Index",
    x = "State",
    y = "Food Environment Index (0-10)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )
```
